Supplementary information - The Functional Therapeutic Chemical Classification System

### 2013-08-15 document.tex
supp figures added
Signed-off-by: loopasam <samuel.croset@gmail.com>

\section{Source code}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the 
FTC can be find at {{https://www.ebi.ac.uk/chembl/ftc}} and the documentation can be 
accessed at {{https://github.com/loopasam/ftc/wiki}}. The reader should be familiar with 
description logics and the Web Ontology Language (OWL) to fully understand the construction of the 
knowledge base. An introduction to description logics from the perspective of the biomedical scientist is 
available on the wiki at {{https://github.com/loopasam/ftc/wiki/Description-Logics}}. The FTC implementation 
relies mostly on Brain (cite) and the web application builds on the top of the Play! framework (cite). 
Classification tasks use the ELK reasoner (cite). The computer hosting the web application has 8Go of memory 
with 4 processors, this architecture allows fast parallel reasoning, thanks to ELK’s design. More functionalities 
will be added to the web application following user requirements (lean implementation).

@@

\subsection{Regulatory pattern}
Some of the FTC categories are created from the biological processes present in the 
GO (cf previous sections); these categories have two arbitrary equivalent definitions, representing the 
two possible ways a compound might impact the biological process. Anti-biological process agent FTC categories 
contain the drugs that negatively perturb a target involved in the positive regulation of the biological process. 
The anti categories also feature the compounds that positively perturb a negative regulator of the same process. 
The pro categories are equivalent to the opposite pattern. Figure 5 illustrates the equivalent definitions for the 
FTC class Anti-blood coagulation agent (FTC_A0007596).

@@

Figure 6: Example of evaluation point. Classes from the ATC are manually 
mapped to classes from the FTC. The drugs present in each side of the equivalences 
are compared, and some metrics are derived from it (false positive/negative and true positive). 
A summary of the evaluation is available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/}}.

@@

\begin{itemize}
\item \textbf{molecular function} \\
Identifier: http://purl.obolibrary.org/obo/GO\_0003674 \\
Label: molecular function \\
Definition: As defined by the Gene Ontology. Elemental activities, such as catalysis or binding, describing 
the actions of a gene product at the molecular level. A given gene product may exhibit one or more molecular functions.
\end{itemize}

\begin{itemize}
\item \textbf{biological process} \\
Identifier: http://purl.obolibrary.org/obo/GO\_0008150 \\
Label: biological process \\
Definition: As defined by the Gene Ontology: Any process specifically pertinent to the functioning of 
integrated living units: cells, tissues, organs, and organisms. A process is a collection of molecular 
events with a defined beginning and end.
\end{itemize}

@@

\begin{itemize}
\item \textbf{Drug} \\
Identifier: http://schema.org/Drug \\
Label: drug \\
Definition: As defined by schema.org. A chemical or biologic substance, used as a medical therapy, that has a 
physiological effect on an organism.
Comment: In the context of the FTC, DrugBank chemicals are considered for their role as therapeutic agent 
rather than for their chemical structure. Mode of action classes are representing agent types and are subclasses of this concept.
\end{itemize}

therapeutic agent
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC_C1
Label: therapeutic agent
Definition: Role of a drug capable of producing a therapeutic effect.

DrugBank compound
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC_C2
Label: DrugBank compound
Definition: Drug coming from DrugBank.

\subsection{Core properties}
The expressivity of the FTC comes mostly for the properties (RBox). Below is a list of the 
OWL object properties present inside the FTC knowledge base.
part-of
Identifier: http://purl.obolibrary.org/obo/BFO_0000050
Characteristic: Transitive
Label: part-of
Definition: As defined and used in the Gene Ontology. More information at http://www.geneontology.org/GO.ontology.relations.shtml#partof
has-part
Identifier: http://purl.obolibrary.org/obo/BFO_0000051
Characteristic: Transitive
Label: has-part
Definition: As defined and used in the Gene Ontology. More information at http://www.geneontology.org/GO.ontology-ext.relations.shtml#haspart
regulates
Identifier: http://purl.obolibrary.org/obo/RO_0002211
Chained property: regulates o part-of -> regulates
Label: regulates
Definition: As defined and used in the Gene Ontology. More information at http://www.geneontology.org/GO.ontology.relations.shtml#regulates
negatively-regulates
Identifier: http://purl.obolibrary.org/obo/RO_0002212
SubPropertyOf: regulates
Label: negatively-regulates
Definition: As defined and used in the Gene Ontology. More information at http://www.geneontology.org/GO.ontology.relations.shtml#regulates
positively-regulates
Identifier: http://purl.obolibrary.org/obo/RO_0002213
SubPropertyOf: regulates
Label: positively-regulates
Definition: As defined and used in the Gene Ontology. More information at http://www.geneontology.org/GO.ontology.relations.shtml#regulates
involved-in
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC_R1
Label: involved-in
Domain: protein
Range: biological process
Definition: Entails the participation of a protein in a biological process
has-function
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC_R2
Label: has-function
Domain: protein
Range: molecular function
Definition: Describes the molecular function a protein can realize.
perturbs
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC_R3
Label: perturbs
Domain: drug
Range: protein
Definition: Specific biochemical interaction through which a drug substance 
will affect the activity of a protein (mechanism of action). The property refers to the specific molecular 
targets to which the drug binds, such as an enzyme or receptor.
negatively-perturbs
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC_R4
Label: negatively-perturbs
SubPropertyOf: perturbs
Definition: Specific biochemical interaction through which a drug substance will decrease the activity of a protein. 
The property refers to the specific molecular targets to which the drug binds, such as an enzyme or receptor.
positively-perturbs
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC_R5
Label: positively-perturbs
SubPropertyOf: perturbs
Definition: Specific biochemical interaction through which a drug substance will increase the activity of a protein. 
The property refers to the specific molecular targets to which the drug binds, such as an enzyme or receptor.

>>>>

\section{Source code}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the 
FTC can be find at {{https://www.ebi.ac.uk/chembl/ftc}} and the documentation can be 
accessed at {{https://github.com/loopasam/ftc/wiki}}. The reader should be familiar with 
description logics and the Web Ontology Language (OWL) to fully understand the construction of the 
knowledge base. An introduction to description logics from the perspective of the biomedical scientist is 
available on the wiki at {{https://github.com/loopasam/ftc/wiki/Description-Logics}}. The FTC implementation 
relies mostly on Brain (cite) and the web application builds on the top of the Play! framework (cite). 
Classification tasks use the ELK reasoner (cite). The computer hosting the web application has 8 Gb of memory 
with 4 processors, this architecture allows fast parallel reasoning, thanks to ELK’s design. More functionalities 
will be added to the web application following user requirements (lean implementation).

@@

\subsection{Regulatory pattern}
Some of the FTC categories are created from the biological processes present in the 
GO (cf previous sections); these categories have two arbitrary equivalent definitions, representing the 
two possible ways a compound might impact the biological process. Anti-biological process agent FTC categories 
contain the drugs that negatively perturb a target involved in the positive regulation of the biological process. 
The anti categories also feature the compounds that positively perturb a negative regulator of the same process. 
The pro categories are equivalent to the opposite pattern. Figure 5 illustrates the equivalent definitions for the 
FTC class Anti-blood coagulation agent (FTC\_A0007596).

@@

\begin{figure}[!tpb]%figureS2
\centerline{\includegraphics{figS2.png}}
\caption{Example of evaluation point. Classes from the ATC are manually 
mapped to classes from the FTC. The drugs present in each side of the equivalences 
are compared, and some metrics are derived from it (false positive/negative and true positive). 
A summary of the evaluation is available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/}}.}\label{fig:S02}
\end{figure}

@@

\begin{itemize}
\item \textbf{Molecular Function} \\
Identifier: http://purl.obolibrary.org/obo/GO\_0003674 \\
Label: molecular function \\
Definition: As defined by the Gene Ontology. Elemental activities, such as catalysis or binding, describing 
the actions of a gene product at the molecular level. A given gene product may exhibit one or more molecular functions.
\end{itemize}

\begin{itemize}
\item \textbf{Biological Process} \\
Identifier: http://purl.obolibrary.org/obo/GO\_0008150 \\
Label: biological process \\
Definition: As defined by the Gene Ontology: Any process specifically pertinent to the functioning of 
integrated living units: cells, tissues, organs, and organisms. A process is a collection of molecular 
events with a defined beginning and end.
\end{itemize}

@@

\begin{itemize}
\item \textbf{Drug} \\
Identifier: http://schema.org/Drug \\
Label: drug \\
Definition: As defined by schema.org. A chemical or biologic substance, used as a medical therapy, that has a 
physiological effect on an organism. \\
Comment: In the context of the FTC, DrugBank chemicals are considered for their role as therapeutic agent 
rather than for their chemical structure. Mode of action classes are representing agent types and are subclasses of this concept.
\end{itemize}

\begin{itemize}
\item \textbf{Therapeutic Agent} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_C1 \\
Label: therapeutic agent \\
Definition: Role of a drug capable of producing a therapeutic effect.
\end{itemize}

\begin{itemize}
\item \textbf{DrugBank Compound} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_C2 \\
Label: DrugBank compound \\
Definition: Drug coming from DrugBank.
\end{itemize}

\subsection{Core properties}
The expressivity of the FTC comes mostly for the properties (RBox). Below is a list of the 
OWL object properties present inside the FTC knowledge base.

\begin{itemize}
\item \textbf{Part Of} \\
Identifier: http://purl.obolibrary.org/obo/BFO\_0000050 \\
Characteristic: Transitive \\
Label: part-of \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Has Part} \\
Identifier: http://purl.obolibrary.org/obo/BFO\_0000051 \\
Characteristic: Transitive \\
Label: has-part \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Regulates} \\
Identifier: http://purl.obolibrary.org/obo/RO\_0002211 \\
Chained property: regulates o part-of -> regulates \\
Label: regulates \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Negatively Regulates} \\
Identifier: http://purl.obolibrary.org/obo/RO\_0002212 \\
SubPropertyOf: regulates \\
Label: negatively-regulates \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Positively Regulates} \\
Identifier: http://purl.obolibrary.org/obo/RO\_0002213 \\
SubPropertyOf: regulates \\
Label: positively-regulates \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Involved In} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_R1 \\
Label: involved-in \\
Domain: protein \\
Range: biological process \\
Definition: Entails the participation of a protein in a biological process. \\
\end{itemize}

\begin{itemize}
\item \textbf{Has Function} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_R2 \\
Label: has-function \\
Domain: protein \\
Range: molecular function \\
Definition: Describes the molecular function a protein can realize. \\
\end{itemize}

\begin{itemize}
\item \textbf{Perturbs} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_R3 \\
Label: perturbs \\
Domain: drug \\
Range: protein \\
Definition: Specific biochemical interaction through which a drug substance 
will affect the activity of a protein (mechanism of action). The property refers to the specific molecular 
targets to which the drug binds, such as an enzyme or receptor.
\end{itemize}

\begin{itemize}
\item \textbf{Negatively Perturbs} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_R4 \\
Label: negatively-perturbs \\
SubPropertyOf: perturbs \\
Definition: Specific biochemical interaction through which a drug substance will decrease the activity of a protein. 
The property refers to the specific molecular targets to which the drug binds, such as an enzyme or receptor.
\end{itemize}

\begin{itemize}
\item \textbf{Positively Perturbs} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_R5 \\
Label: positively-perturbs \\
SubPropertyOf: perturbs \\
Definition: Specific biochemical interaction through which a drug substance will increase the activity of a protein. 
The property refers to the specific molecular targets to which the drug binds, such as an enzyme or receptor.
\end{itemize}

\section*{Pairwise comparison of mode of action similarities}

Supplementary Figure S1: Pairwise comparison of mode of action similarities. 
The compounds are grouped by ATC code (2 levels - legend not shown) on panel A. 
A hierarchical clustering step (based on manhattan distance) was applied on panel B. 
The dendrogram reflects the original structure of the FTC. 
Panel C shows the compounds sorted by DrugBank identifiers only (≈ random). 
The colours displayed on panel B and C correspond to one level ATC categories and are indexed in the legend panel.

\begin{figure*}[!tpb]%figureS3
\centerline{\includegraphics{figS3.png}}
\caption{Caption to be written.}\label{fig:S03}
\end{figure*}
